---
title: "Midlife PA behavioral analysis"
author: "ai.me@northeastern.edu"
output: html_document
date: '2022-04-25'
---

```{r setup, include=FALSE}
library(tidyverse)
library(dplyr)
library(broom)
library(rlang)
library(ggpubr)
library(rstatix)
```
#select the data to use
```{r}
setwd("~/Documents/PAD/PApast/")
df <- read_csv("past_2017_new_all_clean_noClinical.csv")

# cognitive variable kist of the five dimensions
cog <- df %>% select(
  immediate_memory_index_score,
  visuospatial_constructional_index_score,
  language_index_score,
  attention_index_score,
  delayed_memory_index_score
)

# explore relationships between different cognitive measures
pairs.panels(cog, method = "pearson",
             density = TRUE,
             ellipses = TRUE,
             stars=TRUE)

```
#  ########## descriptive data ##########
# demographics
```{r}
#descriptive information of demographic factors
demo <- df %>% summarise(mean_age=mean(age_2017), sd_age=sd(age_2017), mean_edu=mean(Education_years), sd_edu=sd(Education_years))
demo
```
# descriptive visualization about the PA status
```{r}
# plot the four physical status groups separately
MVPA <- df %>% select(ID, exer_past_mvpa, exer_2017_mvpa, act_status) %>% gather(key = time, value = PA, c("exer_past_mvpa","exer_2017_mvpa")) %>% group_by(act_status, time) %>% summarise(MVPA=mean(PA))
MVPA$act_status <- as.factor(MVPA$act_status)
# this is the line plot
MVPA %>% ggplot( aes(x=factor(time, level=c('exer_past_mvpa','exer_2017_mvpa')), y=MVPA, group=act_status, color=act_status)) + xlab(" ")+
    geom_line(size=1.5) + geom_point(size=2.5) + ggtitle("MVPA engagement") + ylab("MVPA engagement (hours/year)")  +scale_color_manual(name="Physical activity status", labels=c("Active stable", 'Active to Inactive','Inactive to Active','Inactive stable'), values=c("salmon","lightskyblue","yellow","lightgreen"))+scale_x_discrete(labels=c("Mid-life MVPA","Later-life MVPA"))
# bar plot
MVPA %>% ggplot( aes(x=factor(time, level=c('exer_past_mvpa','exer_2017_mvpa')), y=MVPA, group=act_status, fill=act_status)) + xlab(" ")+
    geom_bar(stat="identity", position=position_dodge()) + ggtitle("MVPA engagement") + ylab("MVPA engagement (hours/year)")  +scale_fill_manual(name="Physical activity status", labels=c("Active stable", 'Active to Inactive','Inactive to Active','Inactive stable'), values=c("salmon","lightskyblue","yellow","lightgreen"))+scale_x_discrete(labels=c("Mid-life MVPA","Later-life MVPA")) + theme(axis.title.y = element_text(size = 13), axis.text.x=element_text(size=13))
# line plot
TPA <- df %>% select(ID, exer_past_total, exer_2017_total, act_status) %>% gather(key = time, value = PA, c("exer_past_total","exer_2017_total")) %>% group_by(act_status, time) %>% summarise(TPA=mean(PA))
TPA$act_status <- as.factor(TPA$act_status)
TPA %>% ggplot( aes(x=factor(time, level=c('exer_past_total','exer_2017_total')), y=TPA, group=act_status, color=act_status)) +
    geom_line(size=1.5) + geom_point(size=2.5) +ylab("Total PA engagement (hours/year)") + xlab(" ")+  ggtitle("xx") + scale_color_manual(name="Physical activity status", labels=c("Active stable", 'Active to Inactive','Inactive to Active','Inactive stable'), values=c("salmon","lightskyblue","yellow","lightgreen")) + scale_x_discrete(labels=c("Mid-life total PA","Later-life total PA")) + theme(axis.title.y = element_text(size = 13), axis.text.x=element_text(size=13))
# bar plot
TPA %>% ggplot( aes(x=factor(time, level=c('exer_past_total','exer_2017_total')), y=TPA, group=act_status, fill=act_status)) +
    geom_bar(stat="identity", position=position_dodge()) +ggtitle("Total PA engagement") + ylab("Total PA engagement (hours/year)") + xlab(" ")+  ggtitle("xx") + scale_fill_manual(name="Physical activity status", labels=c("Active stable", 'Active to Inactive','Inactive to Active','Inactive stable'), values=c("salmon","lightskyblue","yellow","lightgreen")) + scale_x_discrete(labels=c("Mid-life total PA","Later-life total PA")) + theme(axis.title.y = element_text(size = 13), axis.text.x=element_text(size=13))
past_2017_new_all_clean_noClinical$act_status <- as.factor(past_2017_new_all_clean_noClinical$act_status)


freq <- df[,"act_status"] %>% group_by(act_status) %>% count() %>% ungroup() %>% mutate(perc = `n` / sum(`n`)) %>% 
  arrange(perc) %>%
  mutate(labels = scales::percent(perc))
ggplot(freq, aes(x = "", y = perc, fill=act_status)) +
  geom_col() + geom_text(aes(label=labels), position=position_stack(vjust=0.5))+ coord_polar(theta = "y") + scale_fill_manual(name="Physical activity status", labels=c("Active stable", 'Active to Inactive','Inactive to Active','Inactive stable'), values=c("salmon","lightskyblue","yellow","lightgreen")) + theme_bw() + theme(axis.text = element_blank(),axis.ticks = element_blank(),panel.grid  = element_blank(), axis.title = element_blank())


```
# visualization for the five RBANS scores over 4 timepoints
```{r}
#df_bl4t is the long format of scores across 4 timepoints
dmem_bl4t <- df_bl4t %>% group_by(count) %>% summarise(mean=mean(delayed_memory_index_score, na.rm=TRUE), sd=sd(delayed_memory_index_score, na.rm=TRUE))
imem_bl4t <- df_bl4t %>% group_by(count) %>% summarise(mean=mean(immediate_memory_index_score, na.rm=TRUE), sd=sd(immediate_memory_index_score, na.rm=TRUE))
lang_bl4t <- df_bl4t %>% group_by(count) %>% summarise(mean=mean(language_index_score, na.rm=TRUE), sd=sd(language_index_score, na.rm=TRUE))
vis_bl4t <- df_bl4t %>% group_by(count) %>% summarise(mean=mean(visuospatial_constructional_index_score, na.rm=TRUE), sd=sd(visuospatial_constructional_index_score, na.rm=TRUE))
att_bl4t <- df_bl4t %>% group_by(count) %>% summarise(mean=mean(attention_index_score, na.rm=TRUE), sd=sd(attention_index_score, na.rm=TRUE))
all_bl4t <- df_bl4t %>% group_by(count) %>% summarise(mean=mean(total_scale_index_score, na.rm=TRUE), sd=sd(total_scale_index_score, na.rm=TRUE))

dmem<- ggplot(dmem_bl4t, aes(x=count, y=mean)) + geom_line() + geom_point() + geom_errorbar( aes(ymin = mean-0.1*sd, ymax = mean+0.1*sd),width = 0.2) + xlab("Timepoint") + ylab("Delayed memory")
imem<- ggplot(imem_bl4t, aes(x=count, y=mean)) + geom_line() + geom_point() + geom_errorbar( aes(ymin = mean-0.1*sd, ymax = mean+0.1*sd),width = 0.2) + xlab("Timepoint") + ylab("Immediate memory")
lang<- ggplot(lang_bl4t, aes(x=count, y=mean)) + geom_line() + geom_point() + geom_errorbar( aes(ymin = mean-0.1*sd, ymax = mean+0.1*sd),width = 0.2) + xlab("Timepoint") + ylab("Language")
vis<- ggplot(vis_bl4t, aes(x=count, y=mean)) + geom_line() + geom_point() + geom_errorbar( aes(ymin = mean-0.1*sd, ymax = mean+0.1*sd),width = 0.2) + xlab("Timepoint") + ylab("Visuospatial function")
att<- ggplot(att_bl4t, aes(x=count, y=mean)) + geom_line() + geom_point() + geom_errorbar( aes(ymin = mean-0.1*sd, ymax = mean+0.1*sd),width = 0.2) + xlab("Timepoint") + ylab("Attention")
all<- ggplot(all_bl4t, aes(x=count, y=mean)) + geom_line() + geom_point() + geom_errorbar( aes(ymin = mean-0.1*sd, ymax = mean+0.1*sd),width = 0.2) + xlab("Timepoint") + ylab("Global score")
# plot all at once 
ggarrange(imem,vis,lang,dmem,att, all,ncol=3,nrow=2)
```

# ########## transformation of PA variables #########
```{r, include=FALSE}
library(rcompanion)
set.seed(2004)
# define the function to scale variables to 0-1
scale01 = function(x, ...) {(x - min(x, ...)) / (max(x, ...) - min(x, ...))}

exe_vars <- past_2017_new_all_clean_noClinical %>% dplyr::select(exer_past_total, exer_2017_total, exer_past_mvpa, exer_2017_mvpa, exer_light_past, exer_light_2017 )
exe_vars_trans <- exe_vars %>% lapply(scale01) %>% lapply(transformTukey) 
exe_vars_trans <- do.call(cbind, exe_vars_trans) %>% as_tibble()

write_csv(exe_vars_trans,"exe_vars_trans_156.csv")
```
# compare distribution between not-transformed and transformed
```{r}
#library(data.table)
# library(dplyr)
exe_vars_z <- exe_vars %>% mutate_all(scale)
combine <- list(exe_vars_z, exe_vars_trans) %>% rbindlist(use.names=TRUE, idcol='tran')
```
# plot histograms for trans and non-trans
```{r}
#library(plyr)
#library(ggplot2)
combine$tran <- as.factor(combine$tran)

mu <- ddply(combine, "tran", summarise, grp.mean=mean(exer_2017_mvpa)) # change
ggplot(combine, aes(x=exer_past_total, color=tran, fill=tran)) + # change
geom_histogram(aes(y=..density..), position="identity", alpha=0.5)+
geom_density(alpha=0.6)+
geom_vline(data=mu, aes(xintercept=grp.mean, color=tran),
           linetype="dashed")+
scale_color_manual(values=c("#999999", "#E69F00"), labels=c("Z-transformation","Tukey’s Ladder of Powers"))+
scale_fill_manual(values=c("#999999", "#E69F00"), labels=c("Z-transformation","Tukey’s Ladder of Powers"))+
labs(title="Mid-life total PA",x="Exercise time", y = "Density")+ # change
theme_classic()
```
# plot box plots for transformed scores 
```{r}
ggplot(combine, aes(x=tran, y=exer_past_mvpa, fill=tran)) + scale_fill_manual(values=c("#999999", "#E69F00"), labels=c("Z-transformation","Tukey’s Ladder of Powers"))+
    geom_violin(width=1) + geom_boxplot(width=0.1,outlier.color = "black", outlier.size = 2.5)+coord_flip() + labs(y="Mid-life MVPA", fill="Transformation") + theme(text = element_text(size = 15)) 
```

# ##### Regression ##### 
# diagnose plots
```{r}
diag <- function(y){
  model_str <-paste(y,"~ time*exer_past_total +time*exer_2017_total + time*age_2017+time*CAIDE_sex_subscore + time*CAIDE_APOE4_subscore+time*Education_years")
  par(mfrow = c(2, 2))
  model <- lm(model_str, data = df_4tlong) # long version of RBANS scores across 4 timepoints
  plot(model)
}

lapply(as.list(rbans), diag)
```
# run the mixed linear model
```{r}
library(nlme)
library(lmerTest)

lm_mix <- function(var) {
  var_model<-paste(var, "~count*exer_2017_total+ count*exer_past_total + count*age_2017+count*CAIDE_sex_subscore +count*num_times+ count*CAIDE_APOE4_subscore+count*Education_years + (1|ID)")
  summary(lmerTest::lmer(var_model, REML=FALSE, data=df_bl4t))
}

lm_mix <- function(var) {
  var_model<-paste(var, "~count*act_past*act_curr+count*age_2017+count*CAIDE_sex_subscore + count*num_times +  count*CAIDE_APOE4_subscore+count*Education_years + (1|ID)")
  summary(lmerTest::lmer(var_model, REML=FALSE, data=df_bl4t)) # the interaction analysis
}

rbans=c("attention_index_score", "delayed_memory_index_score", "immediate_memory_index_score", "language_index_score", "total_scale_index_score", "visuospatial_constructional_index_score")
lm_mix_rs <- lapply(as.list(rbans), lm_mix)
```
# FDR
```{r}
#FDR for non-interaction model
p_table<- data.frame(num=c(1,2,3,4,5), pvals_past_total=c(0.08313, 0.12205, 0.09052, 0.6918, 0.00398))#MVPA
p_table<- data.frame(num=c(1,2,3,4,5), pvals_past_total=c(0.81, 0.55, 0.82, 0.32, 0.61))# time*total PA effect
p_table<- data.frame(num=c(1,2,3,4,5), pvals_past_total=c(0.27, 0.01, 0.32, 0.59, 0.01))# time effect total PA
p_table<- data.frame(num=c(1,2,3,4,5), pvals_past_total=c(0.11, 0.15, 0.04, 0.12, 0.08)) # total PA effect
p_table<- data.frame(num=c(1,2,3,4,5), pvals_past_total=c(0.187358, 0.087256, 0.00207, 0.2713, 0.7753)) # mid MVPA effect
p_table<- data.frame(num=c(1,2,3,4,5), pvals_past_total=c(0.37, 0.01, 0.84, 0.53, 0.02)) # time effect MVPA
p_table<- data.frame(num=c(1,2,3,4,5), pvals_past_total=c( 0.28324, 0.16, 0.13, 0.52, 0.32)) # time*MVPA
p_table$fdr <- p.adjust(p_table$pvals, method="fdr")

# FDR for interaction model
p_table<- data.frame(num=c(1,2,3,4,5), pvals=c(0.316, 0.6997, 0.0234, 0.0064, 0.472))#act past 
p_table<- data.frame(num=c(1,2,3,4,5), pvals=c(0.7842, 0.8302, 0.3889, 0.21464, 0.1431))# act curr
p_table<- data.frame(num=c(1,2,3,4,5), pvals=c(0.6038, 0.0722, 0.101, 0.4605, 0.02539))# time 
p_table<- data.frame(num=c(1,2,3,4,5), pvals=c(0.8703, 0.8568, 0.268, 0.0407, 0.96134)) # time and past act
p_table<- data.frame(num=c(1,2,3,4,5), pvals=c(0.081, 0.0755, 0.465, 0.06368, 0.60608)) # time and curr act
p_table<- data.frame(num=c(1,2,3,4,5), pvals=c(0.2832, 0.028, 0.6898, 0.01231, 0.86265)) # interaction
p_table$fdr <- p.adjust(p_table$pvals, method="fdr")
p_table$fdr
```

# ########## interaction plots #########
```{r}
library(car)
library(interactions)
library(jtools)
## Plotting: 3 way time BY midlife PA BY later-life PA
interact_plot(model, pred = count, modx=act_past, mod2=act_curr, x.label = "Timepoints", y.label = "Delayed memory function", legend.main = "Midlife physical activeness",  modx.label = c("Inactive","Active"), mod2.labels = c("Inactive in later-life","Active in later-life"), partial.residuals = TRUE, colors="CUD Bright", interval = TRUE, int.width = 0.9) + theme(axis.title = element_text(size = 14),axis.text=element_text(size=11), legend.text = element_text(size=11),legend.title = element_text(size=11),strip.text = element_text(size=11))


```


