---
title: "RBANS_stats"
author: "ai.me@northeastern.edu"
date: '2023-07-17'
output: html_document
---
# loading required packages
```{r setup, include=FALSE}
library(dplyr)
library(tidyverse)
library(ggplot2)
library(lme4)
library(nlme)
library(lmerTest)
```

# import dataset 
```{r}
#(modify as you need)
```

# preapre your dataset to long version (have one column of timepoints and one participant will have multiple rows)
```{r}
#(modify as you need)
```

# diagnose plots
```{r}
diag <- function(y){
  model_str <-paste(y,"~var1+var2+var3") # this would be the actual model you will be testing
  par(mfrow = c(2, 2))
  model <- lm(model_str, data = df_org_2tlong)
  plot(model)
}

lapply(as.list(cog), diag) #cog would be a vector of cognitive variable names
```

# mixed linear model
```{r}
#this lm_mix function let us loop through multiple cognitive variables
lm_mix <- function(var) {
  var_model<-paste(var, "~var1+var2+ var3 + (1|ID)") #this 1|ID is accounting for individual differences on intercept
  summary(lmerTest::lmer(var_model, REML=FALSE, data=df)) #df would be the long dataset you put together
}

lm_mix_rs <- lapply(as.list(cog), lm_mix) #cog would be a vector of cognitive variable names
#all results will be stored in lm_mix_rs as lists
```
#Note: this function does not automatically give you standardized coefficients; if you need them, scale all the variables before you run the model and the coefficients will be standardized.
